BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-30 ms.tex
Updated draft after editing a few sections

Therefore we introduce \texttt{bhpwave}\footnote{\href{https://bhpwave.readthedocs.io}{https://bhpwave.readthedocs.io}} an open-source Python-based waveform generator that models the adiabatic dynamics and gravitational wave signals produced by a small body undergoing a quasi-circular\footnote{By ``quasi-circular," we mean that, at any moment of time, the inspiral motion is tangent to a circular geodesic as described in Sec.~\ref{sec:adiabatic}. This is a consequence of the fact that in the adiabatic limit, circular orbits remain circular, i.e., $e(t) = O(\epsilon)$ \cite{KennOri96, Kenn98}.} (non-eccentric) inspiral into a {rotating} massive black hole. This code can model any binary with an initial orbital separation of $r_0 \leq 60 M$ and a massive black hole spin in the range $|a| \leq 0.9999 M$. (See Sec.~\ref{sec:adiabatic} for exact definitions of $a$ and $r_0$.) Furthermore, the model supports any mass-ratio, though adiabatic waveform models are most relevant for $\epsilon \lesssim 10^{-4}$. By leveraging parallel computations across CPUs, \texttt{bhpwave} can evaluate years-long waveforms in milliseconds. Even on a standard laptop, waveform evaluations still complete within a few hundred milliseconds to a couple seconds.

@@

Additionally, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined EMRIs for a restricted number massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also included the effects due to the small body's spin \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space. Even when neglecting eccentricity, the inclusion of massive black hole spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors} we use a few example problems test the impact of modeling errors on parameter biases for EMRI data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/amp_combined.pdf}
    \caption{The top panel plots the waveform amplitude $A_{\ell=2, m=2}$ as a function of the radius of the orbit $r_0/M$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The bottom panel plots the $A_{\ell, m=\ell}$ modes as a function of $r_0/M$ for $\hat{a} = 0.9999$ and all $2 \leq \ell \leq 20$. The magnitude of the $\ell = m$ modes is ordered inversely with $\ell$ in the bottom panel. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 20$ being the mode with the smallest magnitude. The bottom (yellow) line in the top panel is identical to the top (yellow) line in the bottom panel.}
    \label{fig:amplitudeVariation}
\end{figure}

>>>>

Therefore we introduce \texttt{bhpwave}\footnote{\href{https://bhpwave.readthedocs.io}{https://bhpwave.readthedocs.io}} an open-source Python-based waveform generator that models the adiabatic dynamics and gravitational wave signals produced by a small body undergoing a quasi-circular\footnote{By ``quasi-circular," we mean that, at any moment of time, the inspiral motion is tangent to a circular geodesic as described in Sec.~\ref{sec:adiabatic}. This is a consequence of the fact that in the adiabatic limit, circular orbits remain circular, i.e., $e(t) = O(\epsilon)$ \cite{KennOri96, Kenn98}.} (non-eccentric, non-precessing) inspiral into a {rotating} massive black hole. This code can model any binary with an initial orbital separation of $r_0 \leq 60 M$ and a massive black hole spin in the range $|a| \leq 0.9999 M$. (See Sec.~\ref{sec:adiabatic} for exact definitions of $a$ and $r_0$.) Furthermore, the model supports any mass-ratio, though adiabatic waveform models are most relevant for $\epsilon \lesssim 10^{-4}$. By leveraging parallel computations across CPUs, \texttt{bhpwave} can evaluate years-long waveforms in milliseconds. Even on a standard laptop, waveform evaluations still complete within a few hundred milliseconds to a couple seconds.

@@

Additionally, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a restricted number massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of massive black hole spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors} we use a few example problems to test the impact of modeling errors on parameter biases for EMRI data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. To support community collaboration and open-science, the codes for calculating all of the data, performing all of the analyses, and generating all of the plots presented in this work are made available via the public Github repository \texttt{bhpwave-article}.\footnote{\href{https://github.com/znasipak/bhpwave-article}{https://github.com/znasipak/bhpwave-article}} For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/amp_combined.pdf}
    \caption{The top panel plots the waveform amplitude $A_{\ell=2, m=2}$ as a function of the radius of the orbit $r_0/M$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The bottom panel plots the $A_{\ell, m=\ell}$ modes as a function of $r_0/M$ for $\hat{a} = 0.9999$ and all $2 \leq \ell \leq 15$. The magnitude of the $\ell = m$ modes is ordered inversely with $\ell$ in the bottom panel. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 15$ being the mode with the smallest magnitude. The bottom (yellow) line in the top panel is identical to the top (yellow) line in the bottom panel.}
    \label{fig:amplitudeVariation}
\end{figure}
